Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-06 method.tex
Minor fixes and change suggestions
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@943 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the compact binary parameter
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
gravitational wave strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' gravitational wave polarizations, so for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\cite{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain (\TD) as an
\fir\ filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point
operations per sample.  However, it is typically much more computationally
efficient to use the convolution theorem and the \fft\ to implement fast
convolution in the frequency domain (\FD), requiring only $\mathcal O(\numtmps
\lg \tmpsamps)$ operations per sample.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.
The reduction in sample count for all of the filters required to search the
entire parameter space is dramatic.

The first transformation is to chop the time-domain templates up into time
slices.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 100$ samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim 100$ times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim
100$.  The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
\emph{time slices} in a time-domain analogue to the frequency-domain
decomposition described in ~\cite{Marion2004, Buskulic2010}.
\editorial{nvf: I've picked the two most relevant; I think they are reasonably complementary.}%
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ sets the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps.  In the next section we
describe how we form our novel detection statistic using the time-slice
decomposition and the \SVD.

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation refering to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation}
and notation refering to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation}

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time-domain or frequency-domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in time slice $s$ \\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

\subsubsection{Extrapolation of the computational cost to an Advanced LIGO search}

>>>>

In this section we describe a decomposition of the \CBC{} waveform
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR2 science runs.
We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
gravitational wave strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' gravitational wave polarizations.  Thus, for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\cite{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time domain (\TD) as an
\fir\ filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point
operations per sample.  However, it is typically much more computationally
efficient to use the convolution theorem and the \fft\ to implement fast
convolution in the frequency domain (\FD), requiring only $\mathcal O(\numtmps
\lg \tmpsamps)$ operations per sample with increased latency.

@@

In the remainder of this section we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.
The reduction in sample count for all of the filters required to search the
entire parameter space is dramatic.
%
\editorial{Drew: is ``sample count" the correct wording?}

The first transformation is to chop the time-domain templates up into time
slices.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim$100.
The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section 
%
\editorial{Drew: We used this wording at the beginning of this paragraph.}
%
we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
\emph{time slices} 
%
\editorial{Drew: This was introduced in the previous section. It should be italicized there.}
%
in a time-domain analogue to the frequency-domain
decomposition described in~\cite{Marion2004, Buskulic2010}.
\editorial{nvf: I've picked the two most relevant; I think they are reasonably complementary.}%
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps.  In the next section we
describe how we form our novel detection statistic using the time-slice
decomposition and the \SVD.

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering waveforms from the the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation refering to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation}
and notation refering to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation}

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time-domain or frequency-domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in time slice $s$ \\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}
\editorial{Drew: Is \slicessamps before or after decimation? We should clarify that here.}

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}
\editorial{Drew:``$1.955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$" can't be true. Should the signs be greater than signs?}

@@

\subsubsection{Extrapolation of the computational cost to an Advanced LIGO search}
\editorial{Drew: I think this section be the last section of the results section. Should we move it to be Sec. 4.2?}

### 2011-06-06 results.tex
Minor fixes and change suggestions
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@943 ccdd05d5-56c1-461a-bf98-85f6876523b4

We expect two known contributions to the \SNR\ loss to arise in our
implementation of the \lloid\ algorithm.  The first is the \SNR\ loss due to
the truncation of the \SVD\ basis and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  The \SNR\ loss is to be compared
with the mismatch of 0.03 that arises from the discreteness of the
template bank which is typically.  We will consider an acceptable
\SNR\ loss to be a factor of 10 smaller than this, i.e., no more than $\sim 0.003$.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined in
~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction of
the physical waveforms one effects directly the mismatch that results from the
truncation of the orthogonal filter matrix.  The conditions presented here are
more complicated than in the original work~\cite{Cannon:2010p10398} due to the
inclusion of the time sliced templates and resampling.  \FIXME{However, we find
that the results are quite as expected.}  Figure \ref{fig:hist-svd-tolerance}
demonstrates that it is possible to acheive typical \SNR\ losses of $\ll1\%$.
However, we do notice that the mismatch saturates with respect to the tolerance.
This could be the effect of the resampling, or another \SNR\ loss that we did
not model or expect.  However the saturation is still an order of magnitude below our target mismatch of $\sim 0.003$.  We find that an \SVD\ tolerance of
$\left(1-10^{-4}\right)$ is adequate. 
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}

>>>>

We expect two known contributions to the \SNR\ loss to arise in our
implementation of the \lloid\ algorithm.  The first is the \SNR\ loss due to
the truncation of the \SVD\ basis and estimates for it
exist~\cite{Cannon:2010p10398}.  The second comes from non ideal
implementations of resampling that cause signal loss.  The \SNR\ loss is to be compared
with the mismatch of 0.03 that arises from the discreteness of the
template bank which is typically.  We will consider an acceptable
\SNR\ loss to be a factor of 10 smaller than this, i.e., no more than $\sim$0.003.

@@

In this section we check the effect of the \SVD\ tolerance parameter defined
in~\cite{Cannon:2010p10398}.  By changing the tolerance of the reconstruction of
the physical waveforms one effects directly the mismatch that results from the
truncation of the orthogonal filter matrix.  The conditions presented here are
more complicated than in the original work~\cite{Cannon:2010p10398} due to the
inclusion of the time sliced templates and resampling.  \FIXME{However, we find
that the results are quite as expected.}  Figure \ref{fig:hist-svd-tolerance}
demonstrates that it is possible to acheive typical \SNR\ losses of $\ll1\%$.
However, we do notice that the mismatch saturates with respect to the tolerance.
This could be the effect of the resampling, or another \SNR\ loss that we did
not model or expect.  However the saturation is still an order of magnitude below our target mismatch of $\sim$0.003.  We find that an \SVD\ tolerance of
$\left(1-10^{-4}\right)$ is adequate. 
%
\begin{figure}
	\begin{center}
		\subfloat[Mismatch versus \textsc{svd} tolerance]{\label{fig:hist-svd-tolerance}\includegraphics{bw.pdf}}
		\subfloat[Mismatch versus $N^\shortuparrow$]{\label{fig:hist-interpolate}\includegraphics{bw_resample.pdf}}
		\caption{Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse respones.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
\subref{fig:hist-svd-tolerance} the interpolation filter length is held fixed
at $N^\shortuparrow = 192$, while the \SVD\ tolerance is varied from
$\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.  In \subref{fig:hist-interpolate}, the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$ is varied from 8 to 192 coefficients.}
	\end{center}
\end{figure}
